% LTeX: language=EN
\section{From High Level to Low Level For Programs}
\label{sec:htl}

\AP In this section, we provide a compilation from \kl{high-level for-programs}
to \kl{simple for-programs}. To smoothen the conversion, we introduce
\intro{generator expressions} to the language, as a way to inline 
function calls. We distinguish between \kl{nested-word} generators $\intro*\ogen{s}$
and boolean generators $\intro*\bgen{s}$.

\paragraph{Generator Expressions.} Let us briefly discuss the new typing rules
and semantics of these \kl{generator expressions}. The meaning of 
$\ogen{s}$ is to evaluate the statement $s$ in the current context and collect its output.
For instance, $\ogen{\hlreturn{x}}$ is equivalent to $x$,
and $\ogen{\hlseq{\hlyield{x}}{\hlyield{y}}}$ is equivalent
to $\olist{x,y}$. Similarly, $\bgen{s}$ is used to evaluate a boolean
statement and return its value. The type of a \kl{generator expression}
is equal to the type of the statement $s$ it contains. Importantly, 
when evaluating the statement $s$ in a generator,
we hide all boolean variables from the \kl{evaluation context}.
In particular, $\hlletboolean{b}{\hlreturn{\bgen{ \hlreturn{b} }}}$ is an \emph{invalid
program}, because the variable $b$ is undefined in the context of the generator
expression $\bgen{b}$. The formal typing rules and semantics of generator expressions 
can be found in \cref{fig:generators}. 

Hiding the booleans from the context, ensures that the evaluation order of 
the expressions is irrelevant, allowing us to freely substitute expressions
during the compilation process. 

\paragraph{Rewriting Steps.} We will convert \kl{for-programs} to
\kl{simple for-programs} by a series of rewriting steps listed below.
While most of the steps make can be applied to any \kl{for-program},
some of them only apply to programs of type $\TOut[1] \to \TOut[1]$.
\begin{enumerate}[label=(\Alph*), ref=Step \Alph*]
    \item \label{item:lit_eq_elim} \intro{Elimination of Literal
        Equalities}, i.e., of expressions $\bliteq{c}{o}$ where $c \in \cexpr$
        and $o \in \oexpr$. This is done by replacing those tests with a call
        to a function that checks for equality with the constant $c$ by
        traversing its input. We define these functions by induction on $c$.
        Note that this is only possible because equalities are always
        between a variable and a constant (\ref{item:equality-checks}).

    \item \label{item:lit_elim} \intro{Elimination of Literal
        Productions}, i.e., of constant expressions in the construction of
        $\oexpr$, except single characters. This is done by replacing a
        constant $c$ by a function call. For instance, $\clist{\cchar{a_1}, \cchar{a_2}}$
        is replaced by a call to a function with body
        $\hlseq{\hlyield{\cchar{a_1}}}{\hlyield{\cchar{a_2}}}$.

    \item \label{item:fun_elim} \intro{Elimination of Function Calls},
        by replacing them with \kl{generator expressions}. Given a function $f$
        with body $s$ and arguments $x_1, \dots, x_n$, we replace a call
        $f(a_1, \dots, a_n)$ by $\ogen{ s[ a_1/x_1, \dots, a_n/x_n ] }$
        (or $\bgen{ \cdots }$ for boolean functions).
        This is valid because functions do not take booleans as arguments
        (\ref{item:boolean-arguments}).

    \item \label{item:bool_elim} \intro{Elimination of Boolean
        Generators}. Note that $\bgen{s}$ can only appear in a conditional test, 
        and let us illustrate this step on an example.
        Consider the following statement:
        $\hlif{\bgen{s_1}}{s_2}{s_3}$. We replace it by $\hlletboolean{b_1}{
        (\hlseq{s_1'}{\hlif{b_1}{s_2}{s_3}}) }$, where $s_1'$ is obtained by
        replacing boolean return statements $(\hlreturn{b})$ by assignments of
        the form $(\hlif{b}{\hlsettrue{b_1}}{\hlskip{}})$.
    \item \label{item:let_output_elim} \intro{Elimination of Let
        Output Statements}, i.e., of statements of the form
        $\hlletoutput{x}{e}{s}$. This is done by textually replacing
        $\hlletoutput{x}{e}{s}$ by $s[x \mapsto e]$.

    \item \label{item:return_elim} \intro{Elimination of Return
        Statements} for \kl{list expressions}. 
        First, to make sure that the program does not produce 
        any output after the first return statement, we
        introduce a boolean variable \texttt{has\_returned}, 
        and guard every yield statement by a check on this variable.
        Then, we replace every statement $\hlreturn{e}$ by
        a for loop $\hlfor{(i,x)}{e}{\hlyield{x}}$.
        This is not possible if the return statement is of type $\TOut[0]$,
        and for this edge case, we refer the readers to our implementation.

    \item \label{item:for_loop_exp} \intro{Expansion of For Loops},
        ensuring that every for loop iterates over a single list
        variable. This is the key step of the compilation, and it will be 
        thoroughly explained later in this section.

    \item \label{item:let_bools_top} 
        \intro{Defining booleans at the beginning of for loops}.
        This is a technical step that ensures that all boolean variables 
        are defined at the beginning of the program or at the beginning of a for loop.
        Thanks to the no-shadowing rule (\ref{item:variable-shadowing}), we can 
        safely move all boolean definitions to the top of their scopes.
\end{enumerate}

\begin{theorem}
    \label{thm:rewriting-termination}
    The rewriting steps (\ref{item:lit_eq_elim}--- \ref{item:let_bools_top})
    all terminate and preserve typing. Moreover, normalized \kl{for-programs}
    of type $ \TOut[1] \to \TOut[1]$ are isomorphic to \kl{simple for-programs}.
\end{theorem}

\paragraph{Forward For Loop Expansion.} We now focus on the \kl{expansion of for
loops}, that is, \ref{item:for_loop_exp}. The case of forward
iterations is simpler and will illustrate a first difficulty. We replace each loop of
the form $\hlfor{(i,x)}{\ogen{s_1}}{s_2}$ by the statement $s_1$ where every
statement $\hlyield{e}$ is replaced by $s_2[x \mapsto e]$. This
rewriting is problematic because it leaves the variable $i$ undefined in $s_2$.
The key observation allowing us to circumvent this issue is that the variable
$i$ can only be used in \emph{comparisons}, and can only be compared with 
variables $j$ that are iterating over $\ogen{s_1}$
(thanks to \ref{item:integer-comparisons}). It is therefore sufficient 
to order the outputs of $s_1$ to effectively remove the variable $i$ from the program.

One can recover the ordering between outputs of $s_1$ by storing
the position of the $\hlyield{e}$ responsible for the output,
together with all position variables visible at that point.
Let us illustrate this in a simple example:
\begin{equation*}
    \hlseq{
    (\hlforRev{(j\tikzmark{yieldIndex},y)}{e}{
        (\hlseq{\tikzmark{yield1}\hlyield{y}}
               {\tikzmark{yield2}\hlyield{\cchar{a}}}
        )})
    }{\tikzmark{yield3}\hlyield{\cchar{b}}}
\end{equation*}
\begin{tikzpicture}[overlay, remember picture]
    % Y1 below = 0.5cm, left 0.3cm of pic cs:yield1
    \node (Y1) at ([yshift=-0.5cm, xshift=0.2cm]pic cs:yield1) {$p_1(j)$};
    \node (Y2) at ([yshift=-0.5cm, xshift=0.2cm]pic cs:yield2) {$p_2(j)$};
    \node (Y3) at ([yshift=-0.5cm, xshift=0.2cm]pic cs:yield3) {$p_3$};
    \node (YI) at ([yshift=0cm, xshift=-0.2cm]pic cs:yieldIndex) {};

    \draw[dashed, A2, thick] (Y1) edge[->, bend left] (YI);
    \draw[dashed, A2, thick] (Y2) edge[->, bend left] (YI);
\end{tikzpicture}
\vspace{1em}

In this example, there are three yield statements at
positions $p_1$, $p_2$ and $p_3$. We can compute
the \emph{happens (strictly) before} relation between outputs 
of the various yield statements:
\begin{center}
    $\mathsf{before}(p_1(j), p_2(j)) = \btrue$ \hspace{1em}
    $\mathsf{before}(p_2(j), p_3) = \btrue$ \hspace{1em}
    $\mathsf{before}(p_1(j), p_3) = \btrue$ 
    \\[1em]
    $\mathsf{before}(p_1(j), p_1(j')) = j > j'$ \hspace{2em}
    $\mathsf{before}(p_2(j), p_2(j')) = j > j'$
    \\[1em]
    $\mathsf{before}(p_1(j), p_2(j')) = j \geq j'$
\end{center}
In the case of $j = j'$, the 
output of $p_1(j)$ happens before the output of $p_2(j')$,
because $p_1$ is the first yield statement in the loop.
When $j > j'$, the output of $p_1(j)$ happens
before the output of $p_2(j')$ because the loop
is iterating in reverse order.

\paragraph{Backward For Loop Expansion.} The case of backward iterations adds a
new layer of complexity, namely to perform a non-reversible computation $s$ in
a reversed order: indeed, in the for loop $\hlforRev{(i,x)}{\ogen{s_1}}{s_2}$,
$s_1$ can contain the command $\hlsettrue{b}$ which cannot be reversed.

Let us consider as an example
$\hlforRev{(i,x)}{\ogen{s}}{\hlyield{x}}$, where the statement $s$ is defined
to print all elements of a list $u$ except the first one, namely:
\[ s \ :=\ \hlletboolean{b}{\hlfor{(j,y)}{ u }{\hlif{b}{\hlyield{y}}{\hlsettrue{b}}}}\] 
The semantics of $\hlforRev{(i,x)}{\ogen{s}}{\hlyield{x}}$ is to print all
elements of $u$ in reverse order, skipping the last loop iteration. To compute
this new statement, we will use the following \emph{trick} that can be traced
back to \cite[Lemma 8.1 and Figure 6, p. 68]{bojanczyk2018polyregular}: we will
use two versions of the statement $s$, the first one $s_\mathsf{rev}$, will be
$s$ where all boolean introductions are removed, if statements
$\hlif{e}{s_1}{s_2}$ are replaced by sequences $\hlseq{s_1}{s_2}$, every loop
direction is swapped, and every sequence of statements is reversed. Its
intended semantics is to reach all possible yield statements of $s$ in the
reversed order. In our case:
\begin{wrapfigure}{r}{0.5\textwidth}
    \centering
\begin{tikzpicture}[overlay, remember picture]
    \draw[fill=D4hint,draw=none] 
                     ([yshift=1em, xshift=-0.5em]pic cs:revloopexpstart) 
                     rectangle 
                     ([yshift=-1.3em, xshift=4.5em]pic cs:revloopexpend);

    \draw[fill=D3hint,draw=none] 
                     ([yshift=1em, xshift=-0.5em]pic cs:revloopexpinnerstart) 
                     rectangle 
                     ([yshift=-0.3em, xshift=4.5em]pic cs:revloopexpend);

    \draw[fill=D2hint,draw=none] 
                     ([yshift=1em, xshift=-0.5em]pic cs:revloopifstart)
                     rectangle 
                     ([yshift=-0.3em, xshift=4.4em]pic cs:revloopifend);

     \node[anchor=east,D4] at ([xshift=1.5em,yshift=2em]pic cs:revloopexpstart) {$s_{\mathsf{rev}}$};
     \node[anchor=west,D3,rotate=90] at ([xshift=11em]pic cs:revloopexpinnerstart) {$s$};
     \node[anchor=west,D2,rotate=90] at ([xshift=9em,yshift=-3em]pic cs:revloopifstart) {yield guard};
\end{tikzpicture}
\begin{align*}
    &\tikzmark{revloopexpstart}\mathsf{for}^{\leftarrow}~(j',y')~\mathsf{in}~ u ~\mathsf{do} \\
    &\quad \tikzmark{revloopexpinnerstart}\mathsf{for}^{\rightarrow}~(j, y)~\mathsf{in}~ u ~\mathsf{do} \\
    &\quad \quad \mathsf{let}~\mathsf{mut}~b = \mathsf{false}~\mathsf{in} \\
    &\quad \quad \mathsf{if} \; b \; \mathsf{then} \\
    &\quad \quad \quad \tikzmark{revloopifstart}\mathsf{if} \; j = j' \; \mathsf{then} \\
    &\quad \quad \quad \quad \mathsf{yield}~y\tikzmark{revloopifend} \\
    &\quad \quad \mathsf{else} \\
    &\quad \quad \quad b \leftarrow \mathsf{true}\tikzmark{revloopexpend} \\
\end{align*}
\caption{Backward for loop expansion.}
\label{fig:revloopexp}
\end{wrapfigure}
\begin{equation*}
    s_\mathsf{rev} \ := \ \hlforRev{(j',y')}{ u }{\hlyield{y'}}
\end{equation*}
Some yield statements are reachable in
$s_\mathsf{rev}$, but not when iterating over $s$ in reverse order.
To ensure that we only output correct elements,
we replace every $\hlyield{\cdot}$ statement in $s_{\mathsf{rev}}$ by a
copy of $s$, leading to the programs $s' = s_{\mathsf{rev}}[ \hlyield{\cdot} \mapsto s ]$.
In our case:
\begin{align*}
    s' \ :=\ \hlforRev{(j',y')}{ u }{ s }
\end{align*}
It is now possible to replace every yield statement in this new program
by a conditional check ensuring that the output would actually be 
produced by the original program $s$.
\begin{equation*}
    s'' = s'[ \hlyield{e} \mapsto \hlifnoelse{\bcomp{i}{=}{j'}}{\hlyield{e}}]
\end{equation*}
In our case, the final program is described in \cref{fig:revloopexp}.
This rewriting can be generalised to any program of the form
$\hlforRev{(i,x)}{\ogen{s_1}}{s_2}$ combining the construction illustrated here
with the one taking care of position variables in the case of forward loops.
